Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2023-10-27 analysis/paper/_results.qmd
round 1 of modifications based on reviewer comments

```{r}
#| label: detection-tbl-setup
uhplc_calculus_detect <- uhplc_calculus_long %>%
#uhplc_calculus_detect <- uhplc_calculus_demography %>%
  #remove_missing() %>%
  ungroup() %>% 
  filter(quant > 0) %>%
  distinct(compound, batch) %>%
  pivot_wider(names_from = batch, values_from = batch) %>% 
  mutate(across(!compound, ~ if_else(!is.na(.x), TRUE, FALSE)))
```

@@

Most plots show a large increase in extracted mass of a compound between the
calculus wash extracts (wash 1-3) and the dissolved calculus (calc). Most samples
containing theophylline and caffeine had the largest quantity of the compound
extracted from the first wash, then decreasing in washes 2 and 3. There is
an increase between wash 3 and the dissolved calculus in all samples.
The patterns are consistent across batches 1 and 2.
Nicotine and cotinine have the same relative quantities in the samples, i.e., the
sample with the highest extracted quantity of nicotine also had the highest extracted
quantity of cotinine [@fig-auth-plot-batch2].

@@

compound_detect_plot <- uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  ggplot(aes(x = compound, fill = batch)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar(position = "dodge") +
    #facet_wrap(~ batch) +
    theme_bw() +
    scale_fill_viridis_d() +
    scale_x_discrete(labels = compound_names) +
    scale_fill_viridis_d(labels = c(1,2)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "left",
      panel.grid.major.x = element_blank()
      ) +
    labs(y = "Count", fill = "Batch")

@@

```{r}
#| label: fig-detection-preservation
#| fig-cap: "(A) Violin plot with overlaid box plots depicting the distribution of extracted quantities of each compound from batch 2 separated by state of preservation of the skeleton. (B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2."
#| fig-width: 5
#| fig-height: 7

preservation_plot <- uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    quant > 0,
    batch == "batch2",
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()

weight_quant_plot <- uhplc_calculus_long %>%
  filter(
    batch == "batch2",
    quant > 0,
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = weight, y = quant)) +
    geom_point(col = viridisLite::viridis(1)) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
    theme_bw() +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    ) +
    labs(y = "Quantity (ng)", x = "Weight (mg)")

@@

```{r}
#| label: tbl-pearson
#| tbl-cap: "Pearson correlation (*r*) on dichotomous skeletal lesions and compound concentrations (ng/mg) from the second batch. Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation. OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid; PN = pipe notches."

@@

corr_tib_out <- corr_tib %>%
  column_to_rownames("var") %>%
  as.matrix() %>% 
  signif(digits = 2)
cols <- 0
for(i in 7:nrow(corr_tib)){ # remove repeated values
  cols <- cols + 1 
  corr_tib_out[i,1:cols] <- ""
} 
corr_names <- c("Caries", "Nicotine", "SA", "Calculus", "PN", "Theophylline", "Caffeine", "Cotinine")
row.names(corr_tib_out) <- c(row.names(corr_tib_out)[1:6], corr_names)
corr_tib_out <- corr_tib_out[-nrow(corr_tib),] # remove last row because it has no values (all duplicate coefficients)
corr_tib_out %>% 
  knitr::kable(col.names = corr_names)
```

@@

corr_tib_out <- corr_tib_long %>%
  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F))

polycorr_tib_out <- polycorr_tib %>%
  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F))
```

@@

```{r}
#| label: fig-polycorr
#| fig-cap: "Plot of the polychoric correlations (*rho*). Larger circles and increased opacity indicates a stronger correlation coefficient. OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl’s nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid."
polycorr$rho %>%
  ggcorrplot::ggcorrplot(
    method = "circle", type = "lower", 
    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::plasma(1, begin = 1)),
    legend.title = expression(rho)
    ) +
  scale_x_discrete(labels = prettified_names) +
  scale_y_discrete(position = "right", labels = prettified_names) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "italic")
  )
```

>>>>

```{r}
#| label: detection-tbl-setup
uhplc_calculus_detect <- uhplc_calculus_long %>%
#uhplc_calculus_detect <- uhplc_calculus_demography %>%
  #remove_missing() %>%
  ungroup() %>% 
  filter(quant > 0) %>%
  distinct(compound, batch) %>%
  pivot_wider(names_from = batch, values_from = batch) %>% 
  mutate(across(!compound, ~ if_else(!is.na(.x), TRUE, FALSE)))

corr_weight_quant <- uhplc_calculus_long %>% 
  ungroup() %>% 
  filter(batch == "batch2",
    quant > 0) %>% 
  select(id, weight, compound, quant) %>% 
  pivot_wider(names_from = compound, values_from = quant) %>% 
  select(where(is.numeric)) %>% 
  cor(use = "pairwise.complete.obs") %>% 
  .[1,]

max_quant <- uhplc_calculus_long %>% 
  filter(batch == "batch2",
    id != "MB329"
    ) %>%
  group_by(compound) %>% 
  summarise(
    max_quant = max(quant, na.rm = T),
    pos_y = 0.8 * max_quant
    )

corr_labels <- tibble(
  compound = names(corr_weight_quant)[-1],
  corr = signif(corr_weight_quant[-1], 2),
) %>% 
  inner_join(max_quant) %>% 
  arrange(compound) 
#%>% 
#  mutate(corr = as.character(corr))
#corr_labels$compounds <- prettified_names[corr_labels$compounds]
```

@@

Most plots show a large increase in extracted mass of a compound between the
calculus wash extracts (wash 1-3) and the dissolved calculus (calc). Most samples
containing theophylline and caffeine had the largest quantity of the compound
extracted from the first wash, then decreasing in washes 2 and 3. There is
an increase between wash 3 and the dissolved calculus in all samples.
The patterns are consistent across batches 1 and 2.
Nicotine and cotinine have the same relative quantities in the samples, i.e., the
sample with the highest extracted quantity of nicotine also had the highest extracted
quantity of cotinine (@fig-auth-plot-batch2).

@@

compound_detect_plot <- uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  ggplot(aes(x = compound, fill = batch)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar(position = "dodge") +
    #facet_wrap(~ batch) +
    theme_bw() +
    #scale_fill_viridis_d() +
    scale_x_discrete(labels = compound_names) +
    scale_fill_viridis_d(labels = c(1,2), end = 0.6) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "left",
      panel.grid.major.x = element_blank()
      ) +
    labs(y = "Count", fill = "Batch")

@@

```{r}
#| label: fig-detection-preservation
#| fig-cap: "(A) Violin plot with overlaid box plots depicting the distribution of extracted quantities of each compound from batch 2 separated by state of preservation of the skeleton. (B) Extracted quantity (ng) of compound plotted against weights of the calculus samples from batch 2. r = Pearson correlation coefficient."
#| fig-width: 4.2
#| fig-asp: 1.6

preservation_plot <- uhplc_calculus_long %>%
  filter(
    !is.na(preservation),
    quant > 0,
    batch == "batch2",
    #id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = preservation, y = quant)) +
    geom_violin(aes(fill = preservation), alpha = 0.6) +
    #geom_jitter(width = 0.2) +
    geom_boxplot(width = 0.2) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
      ) +
    labs(x = "Preservation", y = "Quantity (ng)") +
    scale_fill_viridis_d()

weight_quant_plot <- uhplc_calculus_long %>%
  filter(
    batch == "batch2",
    quant > 0,
    #id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = weight, y = quant)) +
    geom_smooth(method = "lm", se = F, col = viridisLite::viridis(1, begin = 0.6)) +
    geom_point(col = viridisLite::viridis(1)) +
    geom_text(
      data = corr_labels, 
      aes(x = 24, y = pos_y, label = paste("r =", corr)),
      #vjust = -1, hjust = -0.1) +
      ) +

    facet_wrap(
      ~ compound,
      scales = "free_y",
      labeller = labeller(compound = compound_names),
      dir = "v",
      ncol = 1
      ) +

    theme_bw() +
    theme(
     axis.title.y = element_blank(),
     axis.text.y = element_blank(),
     axis.ticks.y = element_blank()
    ) +
    labs(y = "Quantity (ng)", x = "Weight (mg)")

@@

```{r}
#| label: tbl-pearson
#| tbl-cap: "Pearson correlation (*r*) on dichotomous skeletal lesions and compound concentrations (ng/mg) from the second batch. Correlations between pairs of dichotomous variables are removed due to incompatibility with a Pearson correlation. Moderate and strong correlations in **bold**. OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid; PN = pipe notches."

@@

corr_tib_out <- corr_tib %>%
  column_to_rownames("var") %>%
  as.matrix() %>% 
  round(digits = 2)
cols <- 0
for(i in 7:nrow(corr_tib)){ # remove repeated values
  cols <- cols + 1 
  corr_tib_out[i,1:cols] <- ""
} 
corr_names <- c("Caries", "Nicotine", "SA", "Calculus", "PN", "Theophylline", "Caffeine", "Cotinine")
row.names(corr_tib_out) <- c(row.names(corr_tib_out)[1:6], corr_names)
corr_tib_out <- corr_tib_out[-nrow(corr_tib),] # remove last row because it has no values (all duplicate coefficients)
#mode(corr_tib_out) <- "numeric"
options(knitr.kable.NA = "")
corr_tib_out %>% 
  as_tibble(rownames = "var") %>% 
  mutate(across(!var, as.numeric)) %>% 
  tibble::column_to_rownames("var") %>% 
  mutate(across(everything(), ~kableExtra::cell_spec(.x, bold = if_else(.x >= 0.4, T, F)))) %>%   mutate(across(everything(), ~ str_remove_all(.x, "NA"))) %>% 
  #kableExtra::cell_spec(bold = if_else(corr_tib_out >= 0.4, T, F)) #%>% 
  #kableExtra::spec_color(na_color = "fff") %>% 
  knitr::kable(col.names = corr_names)
  #kableExtra::kbl(col.names = corr_names) %>%

@@

corr_tib_out <- corr_tib_long %>%
  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F)) %>% 
  mutate(corr = signif(corr, 2))

polycorr_tib_out <- polycorr_tib %>%
  mutate(across(where(is.character), plyr::revalue, prettified_names_lower, warn_missing = F)) %>% 
  mutate(corr = signif(corr, 2))
```

@@

```{r}
#| label: fig-polycorr
#| fig-cap: "Plot of the polychoric correlations (*rho*). Larger circles and increased opacity indicates a stronger correlation coefficient. OA = osteoarthritis; VOP = vertebral osteophytosis; SN = Schmorl’s nodes; DDD = degenerative disc disease; CO = cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic acid."
#| fig-width: 5
#| fig-height: 4
polycorr$rho %>%
  ggcorrplot::ggcorrplot(
    method = "circle", type = "lower", 
    outline.color = "grey40",
    colors = c(viridisLite::plasma(1, begin = 0), "white", viridisLite::inferno(1, begin = 0.7)),
    legend.title = expression(rho)
    ) +
  scale_x_discrete(labels = prettified_names) +
  scale_y_discrete(position = "right", labels = prettified_names) +
  theme(
    legend.position = "left",
    legend.title = element_text(face = "italic")
  )
```

### 2023-10-27 analysis/paper/paper.qmd
prepare paper for PCI submission
update data source and documentation
tweak first page
round 1 of modifications based on reviewer comments

library(here)
#library(readr)
#devtools::load_all()
source(here("analysis/scripts/setup-qmd.R"))

@@

Keywords: {{< meta keywords >}}

@@

We wish to thank Kirsten Ziesemer for helping track down the calculus samples
from her studies. Additionally, we owe special thanks to Vincent Falger and Kees
de Groot from the Middenbeemster Historical Society for their input on early
draft manuscripts and a lovely guided tour of Middenbeemster.

@@

## Data Availability Statement

All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.7648757>).
Analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium
(<https://doi.org/10.5281/zenodo.7649825>)
using the structure recommended by the **rrtools** R package [@Rrrtools].

>>>>

library(here)
library(knitr)
#library(readr)
#devtools::load_all()
source(here("analysis/scripts/setup-qmd.R"))

@@

<!-- Keywords: {{< meta keywords >}} -->

@@

We wish to thank Kirsten Ziesemer for helping track down the calculus samples
from her studies. Additionally, we owe special thanks to Vincent Falger and Kees
de Groot from the Middenbeemster Historical Society for their input on early
draft manuscripts and a lovely guided tour of Middenbeemster. We also thank
Louise Le Meillour for taking on our preprint and providing helpful comments,
and Mario Zimmerman and two other anonymous reviewers for their time and invaluable
insights.

@@

## Data Availability Statement {-}

All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.8061483>).
Analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium
(<https://doi.org/10.5281/zenodo.7649825>)
using the structure recommended by the **rrtools** R package [@Rrrtools].

## Conflict of interest disclosure {-}

The authors have no conflicts to declare.
